Software Architectures to Integrate Workflow Engines in Science Gateways

### 2016-04-28 paper.tex
Small edits following @merous' corrections.

At the same time, toolboxes were emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular cluster schedulers, to execute tasks at
a high throughput. For instance, FSL can launch tasks on SGE through
its fsl\_sub tool, Nipype has execution plugins for S/OGE, Torque,
IPython, ssh and multi-processor servers, and PSOM is interfaced to
Torque, SGE, MOAB, LSF, and HTCondor. Some of these tools also support
advanced features such as provenance tracking (Nipype, PSOM),
redundancy detection across analyses to avoid re-computation (PSOM),
and so on. A wide-array of workflows have been implemented using these
pipelines and are now shared across neuroimaging groups
world-wide. These domain-specific engines represent a tremendous
opportunity for science gateways to leverage existing tools and
applications. They nicely complement e-Science engines that are more
oriented towards the exploitation of distributed computing
infrastructures, in particular grids and clouds.

For the sake of the analysis conducted in this paper, we define a
\emph{workflow engine} (also abbreviated \emph{engine}) as a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}) and using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow systems and
domain-specific pipelines.\note{Marc-e}{To remain consistent with above descriptions} Some workflow engines, usually e-Science
ones, may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Various science gateways have been developed, including frameworks
used as building blocks to assemble science gateways. The main
frameworks are Apache Airavata~\cite{marru2011apache}, the Catania
Science Gateway Framework~\cite{Ardizzone2012} and
WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateway instances
were built using such
frameworks~\cite{kacsuk2014science,ardizzone2012decide}. Other gateway
instances were also built independently from such specialized framework, using
regular web development tools and libraries to access distributed
systems, for instance CBRAIN~\cite{SHER-14} and VIP~\cite{GLAT-13}.

@@

\paragraph{Extensibility} Integrating a new type of workflow engine
requires to develop interaction \texttt{e} and to install the engine
on the infrastructure (\texttt{E$_1$}=2). Updating an engine version
in this architechture shares the same mechanism as version updates of existing
tasks or tools, which requires an update on the infrastructure (\texttt{E$_2$}=1).  New
workflows are integrated by creating a new task in the science gateway
through interactions \texttt{a$_1$} and \texttt{a$_2$}  (\texttt{E$_3$}=2). Adding a new
infrastructure requires to update interactions \texttt{a$_1$}, \texttt{b} and
\texttt{c} in the science gateway (\texttt{E$_4$}=3).

@@

\paragraph{Global complexity} Setting up a nested workflow
architecture with service invocation requires a science gateway, 2
workflow services and 2 infrastructures (\texttt{I$_1$=5}). The
architecture involves 11 interactions (\texttt{I$_2$=11}):
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c$_1$} (twice: once for each infrastructure),
\texttt{c$_2$} (twice: once for each workflow engine and infrastructure) and
\texttt{d} (twice: once for each workflow engine).
\closedanswerednote{Rafael}{I like the idea of what has been discussed in
  the `integration effort', however I think the name may lead to
  misinterpretations. For example, according to Table 2, it requires
  more effort to integrate in a Tight architecture than in a
  service-based architecture. Considering the principles of software
  engineer, it is simpler to integrate/maintain a system using
  decoupled services. Maybe it should be renamed to something related
  to 'dependencies'.}{Tristan}{In our scope I'm not sure that it's
  simpler to integrate a service architecture than a tightly
  integrated one. But I take your point. 'Dependencies' is not very
  suitable either. What about 'development effort'?}{Tristan}{Replaced
  'Integration effort' with 'global complexity', as discussed in
  github issue \#8.} \note{Marc-e}{I understand both points. However, complexity is
  addressed in Robustness as well... I liked the idea of development effort. It's a tough call.}

>>>>

At the same time, toolboxes were emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular clusters\closednote{Tristan}{Reverted from Marc's 'cluster scheduler'}{}{}, to execute tasks at
a high throughput. For instance, FSL can launch tasks on SGE through
its fsl\_sub tool, Nipype has execution plugins for S/OGE, Torque,
IPython, ssh and multi-processor servers, and PSOM is interfaced to
Torque, SGE, MOAB, LSF, and HTCondor. Some of these tools also support
advanced features such as provenance tracking (Nipype, PSOM),
redundancy detection across analyses to avoid re-computation (PSOM),
and so on. A wide-array of workflows have been implemented using these
pipelines and are now shared across neuroimaging groups
world-wide. These domain-specific engines represent a tremendous
opportunity for science gateways to leverage existing tools and
applications. They nicely complement e-Science engines that are more
oriented towards the exploitation of distributed computing
infrastructures, in particular grids and clouds.

For the sake of the analysis conducted in this paper, we define a
\emph{workflow engine} (also abbreviated \emph{engine}) as a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}) and using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow systems and
domain-specific pipelines.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} Some workflow engines, usually e-Science
ones, may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Various science gateways have been developed, including frameworks
used as building blocks to assemble science gateways. The main
frameworks are Apache Airavata~\cite{marru2011apache}, the Catania
Science Gateway Framework~\cite{Ardizzone2012} and
WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateway instances
were built using such
frameworks~\cite{kacsuk2014science,ardizzone2012decide}. Other gateway
instances were also built independently from such specialized
framework, using regular web development tools and libraries to access
distributed systems. CBRAIN~\cite{SHER-14} and VIP~\cite{GLAT-13} are
two such gateways.

@@

\paragraph{Extensibility} Integrating a new type of workflow engine
requires to develop interaction \texttt{e} and to install the engine
on the infrastructure (\texttt{E$_1$}=2). Updating an engine version
in this architechture shares the same mechanism as version updates of
other tasks, which requires an update on the
infrastructure (\texttt{E$_2$}=1).  New workflows are integrated by
creating a new task in the science gateway through interactions
\texttt{a$_1$} and \texttt{a$_2$} (\texttt{E$_3$}=2). Adding a new
infrastructure requires to update interactions \texttt{a$_1$},
\texttt{b} and \texttt{c} in the science gateway (\texttt{E$_4$}=3).

@@

\paragraph{Global complexity} Setting up a nested workflow
architecture with service invocation requires a science gateway, 2
workflow services and 2 infrastructures (\texttt{I$_1$=5}). The
architecture involves 11 interactions (\texttt{I$_2$=11}):
\texttt{a$_1$} (twice), \texttt{a$_2$}, \texttt{b} (twice), \texttt{c$_1$} (twice),
\texttt{c$_2$} (twice) and
\texttt{d} (twice).
\closedanswerednote{Rafael}{I like the idea of what has been discussed in
  the `integration effort', however I think the name may lead to
  misinterpretations. For example, according to Table 2, it requires
  more effort to integrate in a Tight architecture than in a
  service-based architecture. Considering the principles of software
  engineer, it is simpler to integrate/maintain a system using
  decoupled services. Maybe it should be renamed to something related
  to 'dependencies'.}{Tristan}{In our scope I'm not sure that it's
  simpler to integrate a service architecture than a tightly
  integrated one. But I take your point. 'Dependencies' is not very
  suitable either. What about 'development effort'?}{Tristan}{Replaced
  'Integration effort' with 'global complexity', as discussed in
  github issue \#8.} \note{Marc-e}{I understand both points. However, complexity is
  addressed in Robustness as well... I liked the idea of development effort. It's a tough call.}
